---
title: "Analysis_simulateddataset"
author: "Karim Barigou"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This document demonstrates the simulation of data presented in the course, along with tree-based modeling using the **rpart** package. The analysis includes:

* Simulating the dataset.
* Fitting tree models with different maximum depths.
* Visualizing the tree models.
* Pruning the tree iteratively to identify the optimal model based on cross-validation.

## Setup

We load the required libraries.

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
# Set a seed for reproducibility
set.seed(123)

# Load necessary libraries
library(rpart)
library(rpart.plot)
```

## Data Simulation

We simulate a dataset with 1,000,000 observations. The dataset includes four covariates: Gender (male or female), Age (18 to 65), Split of the annual premium (yes or no), Sports car (yes or no). Then, we compute the expected claim frequency and simulate the number of claims.

```{r data-simulation, echo=TRUE}
# Number of simulations
n <- 1000000

# Simulating the features
X1 <- sample(c("female", "male"), size = n, replace = TRUE, prob = c(0.5, 0.5))
X2 <- sample(18:65, size = n, replace = TRUE, prob = rep(1/48, 48))
X3 <- sample(c("yes", "no"), size = n, replace = TRUE, prob = c(0.5, 0.5))
X4 <- sample(c("yes", "no"), size = n, replace = TRUE, prob = c(0.5, 0.5))

# Compute expected claim frequency μ(x)
mu <- 0.1 * 
  (1 + 0.1 * (X1 == "male")) * 
  (1 + 0.4 * (X2 >= 18 & X2 < 30) + 0.2 * (X2 >= 30 & X2 < 45)) * 
  (1 + 0.15 * (X4 == "yes"))

# Simulate response variable Y (number of claims) following a Poisson distribution
Y <- rpois(n, lambda = mu)

# Combine into a dataframe
data <- data.frame(Y, Gender = X1, Age = X2, Split = X3, Sport = X4)

# Display the first 10 rows of the data
head(data, 10)
```

## Fitting a Tree Model (Maximum Depth = 3)

Here we fit a tree-based model using **rpart** with a maximum tree depth of 3. The model is used to predict the number of claims based on the features. As we have seen in the course, this leads to a **underfitted** model as some important features are missing.

```{r tree-model-depth3, echo=TRUE, fig.height=6, fig.width=8}
# Fit the tree model with maximum depth = 3
tree <- rpart(
  Y ~ Gender + Age + Split + Sport,
  data = data,
  method = "poisson",
  control = rpart.control(minsplit = 50, cp = 0, maxdepth = 3)
)

# Display the tree structure
print(tree)

# Plot the tree
rpart.plot(tree)
```

## Fitting a Full Tree Model (Maximum Depth = 4)

We then fit a full tree model with a maximum depth of 4. As we have seen in the course, this leads to an **overfitted** model given that some unnecessary branches appear now in the tree. 

```{r tree-model-depth4, echo=TRUE, fig.height=6, fig.width=8}
# Fit the tree model with maximum depth = 4
fulltree <- rpart(
  Y ~ Gender + Age + Split + Sport,
  data = data,
  method = "poisson",
  control = rpart.control(cp = 0, maxdepth = 4)
)

# Display the tree structure
print(fulltree)

# Plot the tree
rpart.plot(fulltree)

# Print complexity parameters
printcp(fulltree)
```

## Tree Pruning

Based on the table of the complexity parameters, we know that we will reach the optimal tree (the tree which minimizes the 10-fold cross-validation error) by pruning the tree four times iteratively. To do so, we extract the complexity parameter (cp) table and perform sequential pruning steps.

```{r tree-pruning, echo=TRUE, fig.height=6, fig.width=8}
# Extract the complexity parameter (cp) table
cp_table <- fulltree$cptable

# Get the "before last" cp value and print it
before_last_cp <- cp_table[nrow(cp_table) - 1, "CP"]
print(before_last_cp)

# Prune the tree to the first CP value
tree_pruned_1 <- prune(fulltree, cp = cp_table[nrow(cp_table) - 1, "CP"])
rpart.plot(tree_pruned_1, main = "Pruned Tree - Step 1")

# Prune the tree further using the next CP value
tree_pruned_2 <- prune(tree_pruned_1, cp = cp_table[nrow(cp_table) - 2, "CP"])
rpart.plot(tree_pruned_2, main = "Pruned Tree - Step 2")

# Continue pruning iteratively
tree_pruned_3 <- prune(tree_pruned_2, cp = cp_table[nrow(cp_table) - 3, "CP"])
rpart.plot(tree_pruned_3, main = "Pruned Tree - Step 3")

tree_pruned_4 <- prune(tree_pruned_3, cp = cp_table[nrow(cp_table) - 4, "CP"])
rpart.plot(tree_pruned_4, main = "Pruned Tree - Step 4")

# Print the CP table for the pruned tree, representing the optimal tree based on cross-validation
printcp(tree_pruned_4)
```

